---
title: "Mortality Descriptives"
format:
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
---

## Descriptive Statistics

```{r}
#| echo: false
#| label: load-packages
#| include: false
library(here)
finaldata <- read.csv(here("data","finaldata.csv"))
```

The following document is an exploratory data analysis of global data from the World Bank describing conflict related mortality (i.e., maternal, neonate, under-5, and Infant mortality) between the years 2000-2019.

```{r, include = FALSE}
head(finaldata)
range(finaldata$year)
names(finaldata)
suppressMessages({
  suppressWarnings({
    library(readr)
    library(ggplot2)
    library(dplyr)
  })
})
distinctISO <- finaldata %>%
  summarise(count = n_distinct(ISO))
distinctISO
```

## There are 186 countries represented in this data

The following table shows the countries with the greatest number of annual conflicts between 2000-2019...

```{r fig.height = 10, fig.width = 15,echo = FALSE, warning = FALSE, message = FALSE}
#| echo: false
suppressMessages({
  suppressWarnings({
library(here)
library(readr)
library(ggplot2)
library(table1)
library(knitr)
  })
})
highest_conflicts <- finaldata %>%
  filter(conflict_flag == 1) %>% 
  group_by(ISO) %>%         
  summarise(conflict_count = n(), .groups = 'drop')%>%
  arrange(desc(conflict_count))

kable(highest_conflicts[1:25,],caption="Top 25 countries with the highest number of conflicts")

ggplot(highest_conflicts,
       aes(x = conflict_count,
           y = reorder(ISO, conflict_count)
           )
       ) +
  labs(
    title = "Conflict Count by ISO Code",
    x = "Conflict Count",                      
    y = "ISO Code"                              
  )+
  geom_col()
```

## Now, let us observe the total number of conflict related deaths for each country between 2000-2019...

```{r fig.height = 10, fig.width = 15,echo = FALSE, warning = FALSE, message = FALSE}
#| echo: false
highest_deaths <- finaldata%>%
  group_by(ISO)%>%
  summarise(total_best_sum =sum(total_best, na.rm = TRUE), .groups = 'drop')%>%
  arrange(desc(total_best_sum))
kable(highest_deaths[1:25,],caption="Top 25 countries with the highest number of conflict related deaths")

```

## The top 3 countries with the highest number of conflict related deaths between 2000-2019 are :

1.  **Syria**

2.  **Afghanistan**

3.  **Iraq**

```{r fig.height = 16, fig.width = 15,echo = FALSE, warning = FALSE, message = FALSE}
#| echo: false
highest_deaths%>%
  filter(total_best_sum>=1)%>%
ggplot(aes(x = total_best_sum,
           y = reorder(ISO, total_best_sum)
           )
       )+
  labs(
    title = "Conflict Count by ISO Code",
    x = "Conflict Count",                      
    y = "ISO Code"                              
  ) +
  geom_col()
```

```{r,echo = FALSE, warning = FALSE, message = FALSE, include=FALSE}
#| echo: false
finaldata%>%
summarise(totalsum =sum(total_best, na.rm = TRUE), .groups = 'drop')
  
```

### In total there were 1343366 conflict-related deaths between 2000-2019... Before we analyze other trends, it will be worth it to analyze our data structure and missingness trends

```{r, echo= FALSE, include= FALSE}
str(finaldata)
```

**In total there are 21 variables in our dataset with the following structure:**

-   Character: country_name, iSO, region

-   Integer: year, OECD, OECD2023, Matmor, conflict_flag, drought, earthquake

-   Numeric: gdp1000, popdens, urban, agedep, male_edu, temp, rainfall1000, InfMor, NeoMor, Under5Mor

## Let's check the missingness for all our variables and identify the type of missingness

Let us determine which type of missing data we have in this dataset...

```{r, fig.height = 10, fig.width = 15, echo=FALSE,warning = FALSE, message = FALSE}
library(mice)
library(naniar)
library(dplyr)
mcar_test(finaldata)
```

According to this Missing Completely At Random test, there is sufficient evidence at the 5% level to suggest that our data is not MCAR. This means that our data is either missing at random or missing not at random. Let us take a closer look...

```{r, fig.height = 10, fig.width = 15, include=FALSE, echo=FALSE}

## R MarkdownS
library(mice)
library(dplyr)
library(naniar)
library(foreign)

md.pattern(finaldata[,], rotate.names=TRUE)
```

It is apparent from the plot above that missingness is not a significant concern in our dataset with approximately **87%** not having any missing data across any of the 21 variables. Maternal mortality seems to have the highest level of missingness (413 observations, 12.9%).

```{r, fig.height = 10, fig.width = 30, include= FALSE, echo=FALSE}
library(mice)
library(dplyr)
library(naniar)
library(foreign)
gg_miss_fct(x=finaldata, fct= ISO)
```

From this plot, it is evident that some countries have complete missingness observations for maternal mortality. These countries include:

-   ALB: Albania

-   CIV: Côte d'Ivoire (Ivory Coast)

-   MHL: Marshall Islands

It may be worthwhile to look at the connectedness of missingness across various variables...

```{r, echo=FALSE}
library(mice)
library(dplyr)
library(naniar)
library(foreign)
gg_miss_upset(finaldata, nsets=n_var_miss(finaldata), nintersects= NA)
```

There seems to be two unique trends in missing values across variables— each of which show 18 separate observations of these two trends!

1.  Urban, male_edu, temp, rainfall1000
2.  Infant mortality, neonatal mortality, and under-5 mortality.

# Looking at variable specific trends:

Summarizing the number of conflicts and deaths by region:

```{r, echo=FALSE}
library(knitr)
highest_mregion <- finaldata%>%
  group_by(region)%>%
  summarise(total_region =sum(total_best, na.rm = TRUE), .groups = 'drop')%>%
  arrange(desc(total_region))
kable(highest_mregion[1:17,],caption="Regions with the greatest number of conflict related deaths")

```

It is apparent that African regions have the greatest conflict attributed mortality in the world. Does the same trend occur with the number of conflicts observed between 2000-2019?

```{r, echo=FALSE}
library(knitr)
highest_cregion <- finaldata%>%
  group_by(region)%>%
  summarise(total_cregion =sum(conflict_flag, na.rm = TRUE), .groups = 'drop')%>%
  arrange(desc(total_cregion))
kable(highest_cregion[1:17,],caption="Regions with the greatest number of conflicts in the world")

```

### Let's take a look at more specific types of mortality and relationships of other variables...

```{r, echo=FALSE}
total_dISO <- finaldata%>%
  group_by(ISO)%>%
  summarise(total_dISO =sum(drought, na.rm = TRUE), .groups = 'drop')%>%
  arrange(desc(total_dISO))
kable(total_dISO[1:17,],caption="Countries with the greatest most number of droughts")
```
